---
title: "Intro to _spind_"
author: "Sam Levin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`spind` is a package dedicated to removing the spectre of spatial autocorrelation in your species distribution
models (hereafter referred to as SDMs). It contains many of the tools you need to calculate probabilities of occurence, assess model performance, and conduct multimodel inference for 2-D gridded datasets using methods that are robust to spatial autocorrelation. 

The theory underlying the use of GEEs, WRMs, and many of the other tools in this package is covered elsewhere in the literature, and for the purposes of this vignette, we assume that you have already read those papers. We also assume that you have a working knowledge of how to use _R_. Instead, this vignette will focus on demonstrating how to utilize this package to create an SDM and assess its accuracy. Along the way, we will use a couple different data sets to examine how these functions work and investigate how one might use them to create a robust SDM.

##Generalized estimating equations (GEEs)
Let's start with a fairly simple `GEE` using the simulated `musdata` data set included in the package. 


```{r, echo=FALSE}
library(spind)
```

```{r GEE Data Infiling, eval=FALSE}
data(musdata)
data(carlinadata)

# Examine the structure to familiarize yourself with the data
?musdata
head(musdata)

?carlinadata
head(carlinadata)

```


```{r GEE Example, fig.width=5,fig.height=5}
# Next, fit a simple GEE and view the output
coords<-musdata[ ,4:5]

mgee<-GEE(musculus ~ pollution + exposure, family="poisson", data=musdata,
      coord=coords, corstr="fixed", plot=TRUE, scale.fix=FALSE)

summary(mgee,printAutoCorPars=TRUE)

predictions<-predict(mgee,newdata=musdata)
```

As you can see, this package includes S3 methods for `summary` and `predict`. These are useful in evaluating model fit and autocorrelation of residuals compared to a non-spatial model (in this case, a GLM with the same family as the GEE). Additionally, the `plot` argument in `GEE` can be used to visually inspect the autocorrelation of the residuals from each regression. Note that a QIC (Quasi-information criterion) score is reported as opposed to AIC. This is calculated based on the method described in Hardin & Hilbe (2003) and is implemented using the function `qic.calc`. Please see the references in the documentation of `qic.calc` for more details on how this is calculated.

Note that trying to fit GEEs with `corstr="fixed"` to large data sets will result in errors, as the resulting matrices will be too large to be handled in _R_. This is where fitting clustered models can come in handy. These can be specified by changing the `corstr` to either `"quadratic"` or `"exchangeable"`. See Carl & Kuehn 2007 for more details on how these work. 


## Wavelet Revised Models (WRMs)
Next, we'll examine the other main model that is introduced in this package - the Wavelet Revised Model. Let's start with a fairly simple `WRM` using the same `musdata` data set as above. 

```{r, fig.width=5, fig.height=5}


mwrm<-WRM(musculus ~ pollution + exposure, "poisson", musdata,
coord=coords, level=1, plot=TRUE)

summary(mwrm)

predictions<-predict(mwrm, newdata=musdata)

```

Let's try padding with mean values.

```{r, fig.width=5, fig.height=5}

# Padding with mean values

padded.mwrm<-WRM(musculus ~ pollution + exposure, "poisson", musdata,
                 coord=coords, level=1, pad=list(padform=1), plot=TRUE)

summary(padded.mwrm)

padded.predictions<-predict(padded.mwrm, newdata=musdata)

```

`WRM` has many of the same features as `GEE`. Setting `plot=TRUE` allows you to examine the autocorrelation of residuals from a GLM of the same family as your WRM. S3 methods for `predict` and `summary` allow you to examine outputs from the model using the same code as you might use for a GLM. Also note that this reports an AIC score, rather than a QIC score as in the GEE. 

#### Other features specific to WRMs

`WRM` has a number of other model-specific functions that you may find useful in diagnosing model fit and understanding your results. For example, you might want to plot the variance or covariance of each of your wavelets as a function of `level`. The `covar.plot` function allows you to visually examine the wavelet relationships from your model. However, we are going to switch to the `carlinadata` data set now.

```{r, fig.width=5, fig.height=5}

coords<-carlinadata[ ,4:5]

covar.plot(carlina.horrida ~ aridity + land.use,
           data=carlinadata,coord=coords,wavelet="d4",
           wtrafo='modwt',plot='covar')

covar.plot(carlina.horrida ~ aridity + land.use,
           carlinadata,coord=coords,wavelet="d4",
           wtrafo='modwt',plot='var')

```


## Multi-model inference with GEEs and WRMs
`spind` provides a couple of frameworks for conducting multi-model inference analyses and some helper functions that we hope will make your life easier when examining the results. The first that we'll examine here is the `step.spind` function, which implements step-wise model selection. The process is loosely based on `MASS::stepAIC` and `stats::step`, but is specific to classes `GEE` and `WRM`. Currently, the function only supports backwards model selection. In other words, you have to start with all of the variables in your model formula and remove them in a stepwise fashion. We hope to add forward model selection methods shortly. Additionally, `step.spind` is designed to always respect the heirarchy of variables in the model and the user cannot currently override this. For example, `step.spind` would not remove `age` while retaining `I(age^2)`. However, we are happy to update that feature if a lot of you find it frustrating. We hope you won't :)

We'll go through an example of `step.spind` using a GEE on the `birthwt` data set in the `MASS` package below.

```{r}

library(MASS)
data(birthwt)


x<-rep(1:14,14)
y<-as.integer(gl(14,14))
coords<-cbind(x[-(190:196)],y[-(190:196)])

formula<-formula(low ~ age + lwt + race + smoke + ftv +  bwt)

mgee<-GEE(formula, family = "gaussian", data = birthwt,
          coord=coords, corstr="fixed",scale.fix=TRUE)

ss<-step.spind(mgee,birthwt)

best.mgee<-GEE(ss$model, family = "gaussian", data = birthwt,
           coord=coords, corstr="fixed",scale.fix=TRUE)

summary(best.mgee,printAutoCorPars=FALSE)

```

Additionally, we offer a couple of model selection procedures specific to WRMs. The first is implemented in `mmiWMRR`. This performs a series of scale-specific Wavelet Multi-Resolution Regressions (which are implemented using `scaleWMRR`). It allows one to examine the effect that the `scale` parameter has on the results of their regressions, and then select the appropriate level for subsequent analyses. If you are not already familiar with the `scale` parameter, please see Carl et al (2016) (citation in `scaleWMRR` documentation) before using this function. 

```{r}

data(carlinadata)
coords<- carlinadata[,4:5]

mmi<- mmiWMRR(carlina.horrida ~ aridity + land.use,"poisson",
              carlinadata,coords,scale=3,detail=TRUE,wavelet="d4")

```

One can then take this a step further and visualize these results as a function of scale using the `rvi.plot`. `rvi.plot` uses `mmiWMRR` and repeats the analysis for each level of `scale`, then plots the relative importance of each explanatory variable as function of `scale`. It will also print the resulting model selection tables to the console. 

```{r, fig.width=5, fig.height=5}

rvi.plot(carlina.horrida ~ aridity + land.use,"poisson",
         carlinadata,coords,maxlevel=4,detail=TRUE,wavelet="d4")

```


## Goodness of fit statistics

Once you have your model, whether it be a GEE, WRM, or some other spatial model, you will probably want to look at some other goodness of fit statistics. In this package, these are categorized according to whether or not their values are dependent on the chosen threshold. `th.dep` and `th.indep` are designed to work on any number of model types, all you need is a set of actual values, predictions, and their associated coordinates. We'll use the _hook_ data set to see how these work. 

```{r, eval = FALSE}
data(hook)

# Familiarize yourself with the data
?hook
head(hook)

```


```{r Spatial Indices Example, fig.width=5,fig.height=5,collapes=TRUE,comment="#>"}

df<-hook[,1:2]
coords<-hook[,3:4]

# Threshold dependent metrics
th.dep.indices<-th.dep(data=df,coord=coords,spatial=TRUE)

# Confusion Matrix
th.dep.indices$cm

# Kappa statistic
th.dep.indices$kappa

# Threshold independent metrics
th.indep.indices<-th.indep(data=df,coord=coords,spatial=TRUE,plot.ROC=TRUE)

# AUC
th.indep.indices$AUC

# TSS
th.indep.indices$TSS



```



